Uranyl Solvation by a Three-Dimensional Reference Interaction Site Model

### 2015-06-19 doc/uo2_paper/uo2-II.tex
AM: version sent by BL on Jun 18

%% LyX 2.0.6 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english,journal=jpcafh,manuscript=article,maxauthors=0]{achemso}
\usepackage[T1]{fontenc}
\usepackage{color,soul}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{esint}

@@

\title{\hlc[green]{Uranyl Solvation by a Three-Dimensional Reference Interaction Site
Model}}
\author{Alexei Matveev}
\email{alexei.matveev@gmail.com}
\author{Bo Li}
\author{Notker Rösch}
\email{roesch@mytum.de}
\affiliation{Department Chemie, Technische Universität München, 85747 Garching,
Germany}
\alsoaffiliation{Catalysis Research Center, Technische Universität München, 85747
Garching, Germany}
\alsoaffiliation{Institute of High Performance Computing, Agency for Science, Technology
and Research, 1 Fusionopolis Way, Connexis \#16-16, Singapore 138632,
Singapore}
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\newcommand{\hlc}[2][yellow]{{\sethlcolor{#1}\hl{#2}}}
\usepackage{mathptmx}
\usepackage{times}
%\usepackage{microtype}
%\def\thetable{\arabic{table}}
%\usepackage{ragged2e}
%\usepackage[textfont=it]{caption}
\captionsetup{font={bf}}

@@

The free energy calculations were carried out for the temperature
298~K, with the dielectric constant of bulk water set to 78.4, and
the solvent particle density $\rho=0.0333295$~Å$^{-3}$. These parameters
correspond to a pressure of 1~atm. The 3D domain was represented
by a cubic cell of $20^{3}$~Å$^{3}$ and a Cartesian grid with 96$^{3}$
points and a resolution of about 0.208~Å. The radial grid used to
precompute and interpolate the solvent susceptibility and its contractions
with the Coulomb kernel for use in Eq.~\ref{eq:tau} was chosen with
a radial range of 40~Å and 1536 points corresponding to a resolution
of about 0.026~Å. With these parameters, starting from scratch, a
calculation of the solvent susceptibility, the solvation energy, and
the corresponding forces for uranyl with six explicit water ligands
takes about 40 seconds on 8 cores of two Intel$^{\circledR}$ Xeon$^{\circledR}$
E5630 @ 2.53~GHz CPUs, where one achieves a parallel efficiency above
40\%. \hlc[green]{With the same computing equipment, a comparable
1D-RISM calculation takes about 10 seconds.} The non-linear solver
converges in 8 iterations with about 380 Jacobian applications.

@@

A different source of errors is the approximation of the solute by
a single structure, minimizing the conformation free energy instead
of an ensemble at thermal equilibrium. This approximation can be justified
to some extent for solutes with rigid intra-molecular and strong ligand
bonds, but the weakly associated water molecules of the second solvation
shell do require a statistical treatment. The model with six explicit
water molecules illustrates the point; in that structure, one water
molecule is in the second shell ($n=5+1$, Table~\ref{tab:mm-rism}).
Whereas the structure of the first solvation shell is very similar
for $n=5$ and $n=5+1$ explicit water molecules, the solvation energy,
calculated by Eq.~\ref{eq:dG-sol-definition}, is by about \hlc[green]{5~kcal/mol}
higher (in absolute terms) in the latter case. This is likely an artifact
of the missing statistical averaging over positions and orientations
of the water molecule in the second shell. Hence, in order to yield
numerically stable results, independent of the closure chosen, a successful
model has to treat the strongly bound aqua ligands explicitly as a
part of the solute and weakly associated or free-floating water molecules
as part of the RISM medium.

@@

In QM calculations, we use symmetric models of uranyl aqua complexes
to compare PCM and RISM representations of the solvent in QM calculations.
The under-coordinated species does not feature a vacancy in the first
solvation shell \hlc[green]{without the symmetry constraint}. Instead, all water ligands are equivalent due to
symmetry, also for the five- and six-coordinated uranyl. These models
feature an additional intra-molecular degree of freedom, the uranyl
bond length, in contrast to the MM models discussed in the previous
section where the uranyl bond was fixed at a reasonable value.

When previously exploring these models with 1D RISM \citep{Bo15},
we observed consistently longer U--O and U--O$_{\text{w}}$ bond lengths
in QM+RISM calculations as compared to QM+PCM and experiment. We tentatively
attributed this trend to the excess coordination of uranyl, manifested
by the presence of a peak in the U--O$_{\text{w}}$ RDF at the distance
which corresponds to the first solvation shell for models with explicit
ligands. From the present 3D RISM calculations, it is clear that this
artifact is specific to 1D RISM. Comparing the U--O$_{\text{w}}$
RDFs from the models without any explicit water ligands of uranyl
(Fig.~\ref{fig:rdfs-vs-closure-0w}) and with five such water ligands
(Fig.~\ref{fig:rdfs-vs-closure-5w}), one does not note any such
feature at that crucial separation U--O$_{\text{w}}$. However, the
number integrals, 5.3, 6.2, and 6.4, for the first peak of the U--O$_{\text{w}}$
RDFs with PSE closures of orders 1 to 3, respectively, are overestimated
in the model without explicit aqua ligands (Fig.~\ref{fig:rdfs-vs-closure-0w}).

@@

\caption{Radial distribution functions of solvated uranyl without
  explicit aqua ligands: U--O$_{\text{w}}$ (top panel,
  \hlc[green]{red}), U--H$_{\text{w}}$ (top panel, \hlc[green]{blue}),
  O--O$_{\text{w}}$ (bottom panel, \hlc[green]{red}) and
  O--H$\text{\ensuremath{_{w}}}$ (bottom panel,
  \hlc[green]{blue}). Results obtained with the PSE1 closure are shown
  as \hlc[green]{solid lines}, results obtained with the closure
  relations PSE2 and PSE3 are displayed \hlc[green]{using dashed lines
    and dotted lines respectively}.  \label{fig:rdfs-vs-closure-0w}}

@@

\caption{Potential of mean force, $W(r)$, from the U--O$_{\text{w}}$
  radial distribution function, determined with the closure relations
  PSE1, PSE2 and PSE3, \hlc[green]{are displayed as solid, dashed and
  dot-dashed lines, respectively}. The \hlc[green]{dotted} line
  represents the MD result of Ref.~\citenum{Bo15}.
\label{fig:potential-of-mean-force}}

>>>>

%% LyX 2.0.6 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english,journal=jpcafh,manuscript=article,maxauthors=0]{achemso}
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{esint}

@@

\title{Uranyl Solvation by a Three-Dimensional Reference Interaction Site
Model}
\author{Alexei Matveev}
\email{alexei.matveev@gmail.com}
\author{Bo Li}
\author{Notker Rösch}
\email{roesch@mytum.de}
\affiliation{Department Chemie, Technische Universität München, 85747 Garching,
Germany}
\alsoaffiliation{Catalysis Research Center, Technische Universität München, 85747
Garching, Germany}
\alsoaffiliation{Institute of High Performance Computing, Agency for Science, Technology
and Research, 1 Fusionopolis Way, Connexis \#16-16, Singapore 138632,
Singapore}
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{mathptmx}
\usepackage{times}
%\usepackage{microtype}
%\def\thetable{\arabic{table}}
%\usepackage{ragged2e}
%\usepackage[textfont=it]{caption}
\captionsetup{font={bf}}

@@

The free energy calculations were carried out for the temperature
298~K, with the dielectric constant of bulk water set to 78.4, and
the solvent particle density $\rho=0.0333295$~Å$^{-3}$. These parameters
correspond to a pressure of 1~atm. The 3D domain was represented
by a cubic cell of $20^{3}$~Å$^{3}$ and a Cartesian grid with 96$^{3}$
points and a resolution of about 0.208~Å. The radial grid used to
precompute and interpolate the solvent susceptibility and its contractions
with the Coulomb kernel for use in Eq.~\ref{eq:tau} was chosen with
a radial range of 40~Å and 1536 points corresponding to a resolution
of about 0.026~Å. With these parameters, starting from scratch, a
calculation of the solvent susceptibility, the solvation energy, and
the corresponding forces for uranyl with six explicit water ligands
takes about 40 seconds on 8 cores of two Intel$^{\circledR}$ Xeon$^{\circledR}$
E5630 @ 2.53~GHz CPUs, where one achieves a parallel efficiency above
40\%. With the same computing equipment, a comparable
1D-RISM calculation takes about 10 seconds. The non-linear solver
converges in 8 iterations with about 380 Jacobian applications.

@@

A different source of errors is the approximation of the solute by
a single structure, minimizing the conformation free energy instead
of an ensemble at thermal equilibrium. This approximation can be justified
to some extent for solutes with rigid intra-molecular and strong ligand
bonds, but the weakly associated water molecules of the second solvation
shell do require a statistical treatment. The model with six explicit
water molecules illustrates the point; in that structure, one water
molecule is in the second shell ($n=5+1$, Table~\ref{tab:mm-rism}).
Whereas the structure of the first solvation shell is very similar
for $n=5$ and $n=5+1$ explicit water molecules, the solvation energy,
calculated by Eq.~\ref{eq:dG-sol-definition}, is by about 5~kcal/mol
higher (in absolute terms) in the latter case. This is likely an artifact
of the missing statistical averaging over positions and orientations
of the water molecule in the second shell. Hence, in order to yield
numerically stable results, independent of the closure chosen, a successful
model has to treat the strongly bound aqua ligands explicitly as a
part of the solute and weakly associated or free-floating water molecules
as part of the RISM medium.

@@

In QM+3D-RISM calculations, we use symmetric models of uranyl 
aqua complexes to facilitate a comparison with previous results 
obtained with PCM and 1D-RISM representations of the solvent. 
In these symmetric models all aqua ligands are evenly distributed 
in the equatorial plane. The symmetric structures of the five- and 
six-coordinated uranyl aqua complexes are qualitatively consistent 
with the structures obtained by unconstrained optimization 
(see next section). In contrast, the unconstrained four-coordinated 
species features a vacancy in the first solvation shell, filled by 
the RISM solvent. These QM models feature an additional intra-molecular 
degree of freedom, the uranyl bond length, in contrast to the MM models 
discussed in the previous section where the uranyl bond was fixed at 
a reasonable value.

When previously exploring these symmetric models with 1D RISM \citep{Bo15},
we observed consistently longer U--O and U--O$_{\text{w}}$ bond lengths
in QM+RISM calculations as compared to QM+PCM and experiment. We tentatively
attributed this trend to the excess coordination of uranyl, manifested
by the presence of a peak in the U--O$_{\text{w}}$ RDF at the distance
which corresponds to the first solvation shell for models with explicit
ligands. From the present 3D RISM calculations, it is clear that this
artifact is specific to 1D RISM. Comparing the U--O$_{\text{w}}$
RDFs from the models without any explicit water ligands of uranyl
(Fig.~\ref{fig:rdfs-vs-closure-0w}) and with five such water ligands
(Fig.~\ref{fig:rdfs-vs-closure-5w}), one does not note any such
feature at that crucial separation U--O$_{\text{w}}$. However, the
number integrals, 5.3, 6.2, and 6.4, for the first peak of the U--O$_{\text{w}}$
RDFs with PSE closures of orders 1 to 3, respectively, are overestimated
in the model without explicit aqua ligands (Fig.~\ref{fig:rdfs-vs-closure-0w}).

@@

\caption{Radial distribution functions of solvated uranyl without
  explicit aqua ligands: U--O$_{\text{w}}$ (top panel,
  red), U--H$_{\text{w}}$ (top panel, blue),
  O--O$_{\text{w}}$ (bottom panel, red) and
  O--H$\text{\ensuremath{_{w}}}$ (bottom panel,
  blue). Results obtained with the PSE1 closure are shown
  as solid lines, results obtained with the closure
  relations PSE2 and PSE3 are displayed using dashed lines
    and dotted lines respectively.  \label{fig:rdfs-vs-closure-0w}}

@@

\caption{Potential of mean force, $W(r)$, from the U--O$_{\text{w}}$
  radial distribution function, determined with the closure relations
  PSE1, PSE2 and PSE3, are displayed as solid, dashed and
  dot-dashed lines, respectively. The dotted line
  represents the MD result of Ref.~\citenum{Bo15}.
\label{fig:potential-of-mean-force}}
